---
title: "Plots without Day 12"
author: "heinin"
date: "2025-04-19"
output:
  workflowr::wflow_html:
    code_folding: hide
  html_notebook: default
editor_options:
  chunk_output_type: console
---

## Introduction

Plotting feature expression comparisons etc. with Day 16 samples only.

## Packages and environment variables

```{r, message = F, warning = F, fig.width=12, fig.height=6}

suppressPackageStartupMessages({
  library(Seurat)
  library(SeuratObject)
  library(SeuratDisk)
  library(tidyverse)
  library(tibble)
  library(ggplot2)
  library(ggpubr)
  library(workflowr)
  library(googlesheets4)
  library(CellChat)})

setwd("/home/hnatri/PD1_mm/")
set.seed(9999)
options(ggrepel.max.overlaps = Inf)
options(future.globals.maxSize = 30000 * 1024^2)

# Colors, themes, cell type markers, and plot functions
source("/home/hnatri/PD1_mm/code/utilities.R")
source("/home/hnatri/PD1_mm/code/PD1_mm_themes.R")
source("/home/hnatri/PD1_mm/code/CART_plot_functions.R")

```

## Helper functions

```{r, message = F, warning = F, fig.width=12, fig.height=6}

# Calling DEGs between two groups for each cell type
get_DEGs <- function(seuratdata, celltypes, groupvar, group1, group2){
  DEGlist <- lapply(celltypes, function(xx){
    data_subset <- subset(seuratdata, subset = celltype == xx)
    Idents(data_subset) <- as.character(unlist(data_subset[[groupvar]]))
    
    if (min(table(unlist(data_subset[[groupvar]])))<20){
      return(NULL)
    }
    
    if (all((c(group1, group2) %in% unlist(data_subset[[groupvar]]) == c(T, T)))){
      markers <- FindMarkers(data_subset,
                             ident.1 = group1,
                             ident.2 = group2,
                             assay = "RNA",
                             verbose = F)
      markers$feature <- rownames(markers)
      markers$celltype <- xx
      
      return(markers)
    } else {
      return(NULL)
    }
    })
  
  as.data.frame(do.call(rbind, DEGlist))
}

```

## Import data

```{r, message = F, warning = F, fig.width=12, fig.height=6}

seurat_data <- readRDS("/tgen_labs/banovich/BCTCSF/PD1_mm_Seurat/PD1_mm_Seurat_GSEA.Rds")
DefaultAssay(seurat_data) <- "RNA"

# Updating annotations
gs4_deauth()
markers_annotations  <- gs4_get("https://docs.google.com/spreadsheets/d/1iWYBouwQlQboI-rwiujC0QKJ6lq9XeTffbKm2Nz8es0/edit?usp=sharin#g")
sheet_names(markers_annotations)
annotations <- read_sheet(markers_annotations, sheet = "Cluster annotations")

seurat_data$celltype <- mapvalues(seurat_data$snn_res.0.5,
                                  from = annotations$snn_res.0.5,
                                  to = annotations$annotation)


seurat_data <- subset(seurat_data, subset = Day == 16)

seurat_data$Treatment <- factor(seurat_data$Treatment,
                                levels = c("CTRL", "NEO", "ADJ"))

```

## Feature expression

### Selected lymphoid markers

```{r, message = F, warning = F, fig.width=12, fig.height=16}

lymph_plot_features <- c("Cd3e", "Cd4", "Cd8a", "Tcf7", "Ccr7", "Pdcd1", "Tigit",
                         "Havcr2", "Gzmb", "Gzmk", "Cx3cr1", "Cxcr5", "Foxp3",
                         "Tnfrsf9", "Ifng", "Prf1", "Cdc6", "Lag3", "Klrc1",
                         "Klrg1", "Tox", "Slamf6", "Xcl1", "Mt1", "Nfkbia",
                         "Nr4a1", "Cd69")

lymphoid <- subset(seurat_data, subset = celltype %in% c(grep("^L", unique(seurat_data$celltype), value = T), c("NK", "Treg", "B1", "B2")))
lymphoid$celltype <- factor(lymphoid$celltype,
                            levels = sort(c(grep("^L", unique(lymphoid$celltype), value = T), c("NK", "Treg", "B1", "B2"))))

VlnPlot(lymphoid,
        features = lymph_plot_features,
        group.by = "Treatment",
        cols = treatment_col,
        pt.size = 0,
        ncol = 5,
        slot = "data") &
  theme_classic() &
  NoLegend() &
  ylab("")

```

### DE stats, lineage level

ADJ vs. NEO

```{r, message = F, warning = F, fig.width=10, fig.height=26}

Idents(lymphoid) <- lymphoid$Treatment
    
markers <- FindMarkers(lymphoid,
                       ident.1 = "ADJ",
                       ident.2 = "NEO",
                       assay = "RNA",
                       verbose = F)
markers$feature <- rownames(markers)

markers %>%
  filter(p_val_adj < 0.01,
         abs(avg_log2FC) > 2) %>%
  mutate(p_val = round(p_val, 3),
         p_val_adj = round(p_val_adj, 3))

```

ADJ vs. NEO, T cell markers

```{r, message = F, warning = F, fig.width=10, fig.height=26}

markers %>%
  filter(feature %in% lymph_plot_features) %>%
  mutate(p_val = round(p_val, 3),
         p_val_adj = round(p_val_adj, 3))

```

ADJ vs. CTRL

```{r, message = F, warning = F, fig.width=10, fig.height=26}

Idents(lymphoid) <- lymphoid$Treatment
    
markers <- FindMarkers(lymphoid,
                       ident.1 = "ADJ",
                       ident.2 = "CTRL",
                       assay = "RNA",
                       verbose = F)
markers$feature <- rownames(markers)

markers %>%
  filter(p_val_adj < 0.01,
         abs(avg_log2FC) > 2) %>%
  mutate(p_val = round(p_val, 3),
         p_val_adj = round(p_val_adj, 3))

```

ADJ vs. CTRL, T cell markers

```{r, message = F, warning = F, fig.width=10, fig.height=26}

markers %>%
  filter(feature %in% lymph_plot_features) %>%
  mutate(p_val = round(p_val, 3),
         p_val_adj = round(p_val_adj, 3))

```

NEO vs. CTRL

```{r, message = F, warning = F, fig.width=10, fig.height=26}

Idents(lymphoid) <- lymphoid$Treatment
    
markers <- FindMarkers(lymphoid,
                       ident.1 = "NEO",
                       ident.2 = "CTRL",
                       assay = "RNA",
                       verbose = F)
markers$feature <- rownames(markers)

markers %>%
  filter(p_val_adj < 0.01,
         abs(avg_log2FC) > 2) %>%
  mutate(p_val = round(p_val, 3),
         p_val_adj = round(p_val_adj, 3))

```

NEO vs. CTRL, T cell markers

```{r, message = F, warning = F, fig.width=10, fig.height=26}

markers %>%
  filter(feature %in% lymph_plot_features) %>%
  mutate(p_val = round(p_val, 3),
         p_val_adj = round(p_val_adj, 3))

```

### DE, celltype level

```{r, message = F, warning = F, fig.width=10, fig.height=26}

VlnPlot(lymphoid,
        features = lymph_plot_features,
        group.by = "celltype",
        split.by = "Treatment",
        cols = treatment_col,
        pt.size = 0,
        ncol = 2,
        slot = "data") &
  theme_classic() &
  NoLegend()

```

### Top DE stats

```{r, message = F, warning = F, fig.width=12, fig.height=18}

ADJ_NEO_DEGs <- get_DEGs(seuratdata = lymphoid,
                         celltypes = unique(lymphoid$celltype),
                         groupvar = "Treatment",
                         group1 = "ADJ",
                         group2 = "NEO")

ADJ_CTRL_DEGs <- get_DEGs(seuratdata = lymphoid,
                          celltypes = unique(lymphoid$celltype),
                          groupvar = "Treatment",
                          group1 = "ADJ",
                          group2 = "CTRL")

NEO_CTRL_DEGs <- get_DEGs(seuratdata = lymphoid,
                          celltypes = unique(lymphoid$celltype),
                          groupvar = "Treatment",
                          group1 = "NEO",
                          group2 = "CTRL")

ADJ_NEO_DEGs_sig <- ADJ_NEO_DEGs %>%
  filter(p_val_adj < 0.01,
         abs(avg_log2FC) > 2) %>%
  mutate(p_val = round(p_val, 3),
         p_val_adj = round(p_val_adj, 3))

ADJ_CTRL_DEGs_sig <- ADJ_CTRL_DEGs %>%
  filter(p_val_adj < 0.01,
         abs(avg_log2FC) > 2) %>%
  mutate(p_val = round(p_val, 3),
         p_val_adj = round(p_val_adj, 3))

NEO_CTRL_DEGs_sig <- NEO_CTRL_DEGs %>%
  filter(p_val_adj < 0.01,
         abs(avg_log2FC) > 2) %>%
  mutate(p_val = round(p_val, 3),
         p_val_adj = round(p_val_adj, 3))

```

ADJ_NEO_DEGs_sig

```{r, message = F, warning = F, fig.width=12, fig.height=18}

ADJ_NEO_DEGs_sig

```

ADJ_CTRL_DEGs_sig

```{r, message = F, warning = F, fig.width=12, fig.height=18}

ADJ_CTRL_DEGs_sig

```

NEO_CTRL_DEGs_sig

```{r, message = F, warning = F, fig.width=12, fig.height=18}

NEO_CTRL_DEGs_sig

```

### T cell marker DE stats

ADJ vs. NEO

```{r, message = F, warning = F, fig.width=12, fig.height=18}

ADJ_NEO_DEGs %>%
  filter(feature %in% lymph_plot_features) %>%
  mutate(p_val = round(p_val, 3),
         p_val_adj = round(p_val_adj, 3))

```

ADJ vs. CTRL

```{r, message = F, warning = F, fig.width=12, fig.height=18}

ADJ_CTRL_DEGs %>%
  filter(feature %in% lymph_plot_features) %>%
  mutate(p_val = round(p_val, 3),
         p_val_adj = round(p_val_adj, 3))

```

NEO vs. CTRL

```{r, message = F, warning = F, fig.width=12, fig.height=18}

NEO_CTRL_DEGs %>%
  filter(feature %in% lymph_plot_features) %>%
  mutate(p_val = round(p_val, 3),
         p_val_adj = round(p_val_adj, 3))

```

### Selected myeloid markers

```{r, message = F, warning = F, fig.width=12, fig.height=18}

# Mr1, H2-genes, 
#grep("^H2-", rownames(seurat_data), value = T)
#grep("^Apoe", rownames(seurat_data), value = T)

myel_plot_features <- c("Cd68", "Spp1", "Mr1", "C1qa", "C1qb", "C1qc", "Apoe",
                         grep("^H2-", rownames(seurat_data), value = T))

myeloid <- subset(seurat_data, subset = celltype %in% c(grep("^M", unique(seurat_data$celltype), value = T), c("Neut1")))
myeloid$celltype <- factor(myeloid$celltype,
                           levels = sort(c(grep("^M", unique(myeloid$celltype), value = T), c("Neut1"))))

VlnPlot(myeloid,
        features = myel_plot_features,
        group.by = "Treatment",
        cols = treatment_col,
        pt.size = 0,
        slot = "data") &
  theme_classic() &
  NoLegend() &
  ylab("")

```

```{r, message = F, warning = F, fig.width=10, fig.height=28}

VlnPlot(myeloid,
        features = myel_plot_features,
        group.by = "celltype",
        split.by = "Treatment",
        cols = treatment_col,
        pt.size = 0,
        ncol = 2,
        slot = "data") &
  theme_classic() &
  NoLegend()

```



