---
title: "Comparative analysis of the Kluc tumor scRNAseq data"
author: "heinin"
date: "2024-02-26"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

Comparing treatment groups/timepoints.

### Packages and environment variables

```{r, warning=F, message=F}

suppressPackageStartupMessages({
  #library(cli)
  library(Seurat)
  library(SeuratObject)
  library(SeuratDisk)
  library(tidyverse)
  library(tibble)
  library(ggplot2)
  library(ggpubr)
  library(ggrepel)
  library(workflowr)
  library(googlesheets4)
  library(scProportionTest)
  library(UpSetR)})

setwd("/home/hnatri/PD1_mm/")
set.seed(9999)
options(ggrepel.max.overlaps = Inf)

# Colors, themes, cell type markers, and plot functions
source("/home/hnatri/PD1_mm/code/utilities.R")
source("/home/hnatri/PD1_mm/code/PD1_mm_themes.R")
source("/home/hnatri/PD1_mm/code/CART_plot_functions.R")

```

### Importing data

```{r}

seurat_data <- readRDS("/tgen_labs/banovich/BCTCSF/PD1_mm_Seurat/PD1_mm_Seurat_merged.Rds")
seurat_data$Treatment_Day <- paste0(seurat_data$Treatment, "_", seurat_data$Day)

```

### DimPlot of cell type annotations

```{r}

celltypes <- sort(as.character(unique(seurat_data$celltype)))
plot_colors <- colorRampPalette(brewer.pal(11, "Paired"))(length(celltypes))
names(plot_colors) <- celltypes

DimPlot(seurat_data,
        group.by = "celltype",
        reduction = "umap",
        raster = T,
        label = T,
        cols = plot_colors) &
  coord_fixed(ratio = 1) &
  theme_bw() &
  NoLegend()

DimPlot(seurat_data,
        group.by = "celltype",
        split.by = "Treatment",
        reduction = "umap",
        raster = T,
        #label = T,
        cols = plot_colors) &
  coord_fixed(ratio = 1) &
  theme_bw() &
  NoLegend()

```

### Cell type proportion differences

In scatter plots, the proportions of cell types in each pair of treatment groups
are plotted against each other with one group on each axis. The forest plots show
the significance level.

```{r}

unique(seurat_data$Treatment)
unique(seurat_data$Day)

table(seurat_data$celltype, seurat_data$Treatment)

create_clusterpropplot(seurat_data,
                       group_var = "Treatment",
                       group1 = "ADJ",
                       group2 = "CTRL",
                       plot_var = "celltype",
                       plot_colors = plot_colors,
                       var_names = c("ADJ", "CTRL"),
                       legend_title = "Treatment")

create_clusterpropplot(seurat_data,
                       group_var = "Treatment",
                       group1 = "NEO",
                       group2 = "CTRL",
                       plot_var = "celltype",
                       plot_colors = plot_colors,
                       var_names = c("NEO", "CTRL"),
                       legend_title = "Treatment")

create_clusterpropplot(seurat_data,
                       group_var = "Treatment",
                       group1 = "ADJ",
                       group2 = "NEO",
                       plot_var = "celltype",
                       plot_colors = plot_colors,
                       var_names = c("ADJ", "NEO"),
                       legend_title = "Treatment")

# Day 12
create_clusterpropplot(subset(seurat_data, subset = Day == 12),
                       group_var = "Treatment",
                       group1 = "NEO",
                       group2 = "CTRL",
                       plot_var = "celltype",
                       plot_colors = plot_colors,
                       var_names = c("NEO", "CTRL"),
                       legend_title = "Treatment, Day 12")

# Day 16
create_clusterpropplot(subset(seurat_data, subset = Day == 16),
                       group_var = "Treatment",
                       group1 = "ADJ",
                       group2 = "CTRL",
                       plot_var = "celltype",
                       plot_colors = plot_colors,
                       var_names = c("ADJ", "CTRL"),
                       legend_title = "Treatment, Day 16")

create_clusterpropplot(subset(seurat_data, subset = Day == 16),
                       group_var = "Treatment",
                       group1 = "NEO",
                       group2 = "CTRL",
                       plot_var = "celltype",
                       plot_colors = plot_colors,
                       var_names = c("NEO", "CTRL"),
                       legend_title = "Treatment, Day 16")

create_clusterpropplot(subset(seurat_data, subset = Day == 16),
                       group_var = "Treatment",
                       group1 = "ADJ",
                       group2 = "NEO",
                       plot_var = "celltype",
                       plot_colors = plot_colors,
                       var_names = c("ADJ", "NEO"),
                       legend_title = "Treatment, Day 16")


# Using scProportionTest
prop_test <- sc_utils(seurat_data)

prop_test <- permutation_test(
  prop_test, cluster_identity = "celltype",
  sample_1 = "CTRL", sample_2 = "ADJ",
  sample_identity = "Treatment")

perm_plot <- permutation_plot(prop_test)

perm_plot + scale_colour_manual(values = c("tomato", "azure2")) +
  #NoLegend() +
  ggtitle("ADJ vs. CTRL, all timepoints")

prop_test <- permutation_test(
  prop_test, cluster_identity = "celltype",
  sample_1 = "CTRL", sample_2 = "NEO",
  sample_identity = "Treatment")

perm_plot <- permutation_plot(prop_test)

perm_plot + scale_colour_manual(values = c("tomato", "azure2")) +
  #NoLegend() +
  ggtitle("NEO vs. CTRL, all timepoints")

prop_test <- permutation_test(
  prop_test, cluster_identity = "celltype",
  sample_1 = "ADJ", sample_2 = "NEO",
  sample_identity = "Treatment")

perm_plot <- permutation_plot(prop_test)

perm_plot + scale_colour_manual(values = c("tomato", "azure2")) +
  #NoLegend() +
  ggtitle("NEO vs. ADJ, all timepoints")

# Day 12 only
# For day 12, no ADJ sample
prop_test <- sc_utils(subset(seurat_data, subset = Day == 12))

prop_test <- permutation_test(
  prop_test, cluster_identity = "celltype",
  sample_1 = "CTRL", sample_2 = "NEO",
  sample_identity = "Treatment")

perm_plot <- permutation_plot(prop_test)

perm_plot + scale_colour_manual(values = c("tomato", "azure2")) +
  #NoLegend() +
  ggtitle("NEO vs. CTRL, day 12")

# Day 16 only
prop_test <- sc_utils(subset(seurat_data, subset = Day == 16))

prop_test <- permutation_test(
  prop_test, cluster_identity = "celltype",
  sample_1 = "CTRL", sample_2 = "ADJ",
  sample_identity = "Treatment")

perm_plot <- permutation_plot(prop_test)

perm_plot + scale_colour_manual(values = c("tomato", "azure2")) +
  #NoLegend() +
  ggtitle("ADJ vs. CTRL, day 16")

prop_test <- permutation_test(
  prop_test, cluster_identity = "celltype",
  sample_1 = "CTRL", sample_2 = "NEO",
  sample_identity = "Treatment")

perm_plot <- permutation_plot(prop_test)

perm_plot + scale_colour_manual(values = c("tomato", "azure2")) +
  #NoLegend() +
  ggtitle("NEO vs. CTRL, day 16")

prop_test <- permutation_test(
  prop_test, cluster_identity = "celltype",
  sample_1 = "ADJ", sample_2 = "NEO",
  sample_identity = "Treatment")

perm_plot <- permutation_plot(prop_test)

perm_plot + scale_colour_manual(values = c("tomato", "azure2")) +
  #NoLegend() +
  ggtitle("NEO vs. ADJ, day 16")

```

### Cluster markers

```{r}

# Dropping MT and RP genes before calling markers
RBMTgenes <- grep(pattern = "^RP[SL]|^MRP[SL]|^MT-",
                  x = rownames(seurat_data@assays$RNA_human@data),
                  value = TRUE, invert = TRUE)
seurat_data <- subset(seurat_data, features = RBMTgenes)

# Top markers for each cluster
markers <- presto::wilcoxauc(seurat_data,
                             group_by = "celltype",
                             assay = "data",
                             seurat_assay = "RNA_human")

sig_markers <- markers %>% filter(auc>0.8, padj<0.01)

dim(sig_markers)
table(sig_markers$group)

table(sig_markers$group) %>% as.data.frame() %>%
  ggplot(aes(x = reorder(Var1, -Freq), y = Freq, fill = Var1)) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = plot_colors) +
    theme_classic2() +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
    NoLegend() +
    xlab("Cell type") +
    ylab("# cell type markers")

# Saving to a file
write.table(sig_markers, "/scratch/hnatri/CART/PD1_CART_celltype_markers.tsv",
            sep = "\t", quote = F, row.names = F)

# Top markers for each cluster
top_markers <- markers %>% group_by(group) %>% slice_max(order_by = auc, n = 20)

write.table(top_markers, "/home/hnatri/PD1_mm/PD1_CART_celltype_markers_top20.tsv",
            quote = F, row.names = F, sep = "\t")

```

### DEGs between treatment groups

```{r}

# For each cluster, top DEGs between NEO and CTRL
DEG_NEO_CTRL <- lapply(unique(seurat_data$celltype), function(xx){
  data_subset <- subset(seurat_data, subset = celltype == xx)
  if (all((c("NEO", "CTRL") %in% data_subset$Treatment) == c(T, T))){
    markers <- presto::wilcoxauc(data_subset,
                                 group_by = "Treatment",
                                 groups_use = c("NEO", "CTRL"),
                                 assay = "data",
                                 seurat_assay = "RNA_human")
    markers$celltype <- xx
    
    return(markers)
  } else {
    return(NULL)
  }
})
names(DEG_NEO_CTRL) <- unique(seurat_data$celltype)
DEG_NEO_CTRL[sapply(DEG_NEO_CTRL, is.null)] <- NULL

DEG_NEO_CTRL_df <- as.data.frame(do.call(rbind, DEG_NEO_CTRL))

DEG_NEO_CTRL_df_sig <- DEG_NEO_CTRL_df %>%
  filter(group=="NEO",
         padj < 0.01,
         abs(logFC) > 10,
         (pct_in > 50 | pct_out > 50))

# Saving to a file
write.table(DEG_NEO_CTRL_df_sig,
            "/scratch/hnatri/CART/PD1_CART_DEG_NEO_CTRL_sig.tsv",
            sep = "\t", quote = F, row.names = F)

# Comparing to the Seurat function
DEG_NEO_CTRL_Seurat <- lapply(unique(seurat_data$celltype), function(xx){
  data_subset <- subset(seurat_data, subset = celltype == xx)
  Idents(data_subset) <- data_subset$Treatment
  if (all((c("NEO", "CTRL") %in% data_subset$Treatment) == c(T, T))){
    markers <- FindMarkers(data_subset,
                           ident.1 = "NEO",
                           ident.2 = "CTRL",
                           assay = "RNA_human",
                           verbose = F)
    markers$feature <- rownames(markers)
    markers$celltype <- xx
    
    return(markers)
  } else {
    return(NULL)
  }
})
names(DEG_NEO_CTRL_Seurat) <- unique(seurat_data$celltype)
DEG_NEO_CTRL_Seurat[sapply(DEG_NEO_CTRL_Seurat, is.null)] <- NULL

DEG_NEO_CTRL_Seurat_df <- as.data.frame(do.call(rbind, DEG_NEO_CTRL_Seurat))

# Distribution of log2FC
hist(DEG_NEO_CTRL_Seurat_df$avg_log2FC, breaks = 100)

DEG_NEO_CTRL_Seurat_df_sig <- DEG_NEO_CTRL_Seurat_df %>%
  filter(p_val_adj < 0.01,
         abs(avg_log2FC) > 100,
         (pct.1 > 0.50 | pct.2 > 0.50))

# Saving to a file
write.table(DEG_NEO_CTRL_Seurat_df_sig,
            "/scratch/hnatri/CART/PD1_CART_DEG_NEO_CTRL_Seurat_sig.tsv",
            sep = "\t", quote = F, row.names = F)

# DEGs shared between and unique to each method
length(intersect(DEG_NEO_CTRL_Seurat_df_sig$feature, DEG_NEO_CTRL_df_sig$feature))
length(setdiff(DEG_NEO_CTRL_Seurat_df_sig$feature, DEG_NEO_CTRL_df_sig$feature))
length(setdiff(DEG_NEO_CTRL_df_sig$feature, DEG_NEO_CTRL_Seurat_df_sig$feature))

hist(DEG_NEO_CTRL_Seurat_df_sig$avg_log2FC)
hist(DEG_NEO_CTRL_df_sig$logFC)

hist(DEG_NEO_CTRL_Seurat_df_sig$p_val_adj)
hist(DEG_NEO_CTRL_df_sig$padj)

# Comparing logFC values
merge(DEG_NEO_CTRL_Seurat_df, DEG_NEO_CTRL_df, by = c("feature", "celltype")) %>%
  ggplot(aes(x = avg_log2FC, y = logFC)) +
    geom_point() +
    theme_classic2()

# Comparing logFC values for significant DEGs only
merge(DEG_NEO_CTRL_Seurat_df_sig, DEG_NEO_CTRL_df_sig, by = c("feature", "celltype")) %>%
  ggplot(aes(x = avg_log2FC, y = logFC)) +
    geom_point() +
    theme_classic2()

# Comparing adjusted p-values
merge(DEG_NEO_CTRL_Seurat_df, DEG_NEO_CTRL_df, by = c("feature", "celltype")) %>%
  ggplot(aes(x = p_val_adj, y = padj)) +
    geom_point() +
    theme_classic2()

# Comparing adjusted p-values for significant DEGs
merge(DEG_NEO_CTRL_Seurat_df_sig, DEG_NEO_CTRL_df_sig, by = c("feature", "celltype")) %>%
  ggplot(aes(x = p_val_adj, y = padj)) +
    geom_point() +
    theme_classic2()

# Plotting numbers of DEGs
table(DEG_NEO_CTRL_df_sig$celltype) %>% as.data.frame() %>%
  ggplot(aes(x = reorder(Var1, -Freq), y = Freq, fill = Var1)) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = plot_colors) +
    theme_classic2() +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
    NoLegend() +
    xlab("Cell type") +
    ylab("# DEGs, NEO vs. CTRL, all timepoints")

table(DEG_NEO_CTRL_Seurat_df_sig$celltype) %>% as.data.frame() %>%
  ggplot(aes(x = reorder(Var1, -Freq), y = Freq, fill = Var1)) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = plot_colors) +
    theme_classic2() +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
    NoLegend() +
    xlab("Cell type") +
    ylab("# DEGs, NEO vs. CTRL, all timepoints, Seurat")

# Top DEGs for each cluster
DEG_NEO_CTRL_df_sig %>% group_by(group) %>% slice_max(order_by = auc, n = 10)
DEG_NEO_CTRL_Seurat_df_sig %>% group_by(celltype) %>% slice_max(order_by = avg_log2FC, n = 10)

# For each cluster, top DEGs between NEO and ADJ
DEG_NEO_ADJ <- lapply(unique(seurat_data$celltype), function(xx){
  data_subset <- subset(seurat_data, subset = celltype == xx)
  if (all((c("NEO", "ADJ") %in% data_subset$Treatment) == c(T, T))){
    markers <- presto::wilcoxauc(data_subset,
                                 group_by = "Treatment",
                                 groups_use = c("NEO", "ADJ"),
                                 assay = "data",
                                 seurat_assay = "RNA_human")
    markers$celltype <- xx
    
    return(markers)
  } else {
    return(NULL)
  }
})
names(DEG_NEO_ADJ) <- unique(seurat_data$celltype)
DEG_NEO_ADJ[sapply(DEG_NEO_ADJ, is.null)] <- NULL

DEG_NEO_ADJ_df <- as.data.frame(do.call(rbind, DEG_NEO_ADJ))

DEG_NEO_ADJ_df_sig <- DEG_NEO_ADJ_df %>%
  filter(group == "NEO",
         padj < 0.01,
         auc > 0.6,
         (pct_in > 50 | pct_out > 50))

# Saving to a file
write.table(DEG_NEO_ADJ_df_sig,
            "/scratch/hnatri/CART/PD1_CART_DEG_NEO_ADJ_sig.tsv",
            sep = "\t", quote = F, row.names = F)

# Calling DEGs with Seurat
DEG_NEO_ADJ_Seurat <- lapply(unique(seurat_data$celltype), function(xx){
  data_subset <- subset(seurat_data, subset = celltype == xx)
  Idents(data_subset) <- data_subset$Treatment
  if (all((c("NEO", "ADJ") %in% data_subset$Treatment) == c(T, T))){
    markers <- FindMarkers(data_subset,
                           ident.1 = "NEO",
                           ident.2 = "ADJ",
                           assay = "RNA_human",
                           verbose = F)
    markers$feature <- rownames(markers)
    markers$celltype <- xx
    
    return(markers)
  } else {
    return(NULL)
  }
})
names(DEG_NEO_ADJ_Seurat) <- unique(seurat_data$celltype)
DEG_NEO_ADJ_Seurat[sapply(DEG_NEO_ADJ_Seurat, is.null)] <- NULL

DEG_NEO_ADJ_Seurat_df <- as.data.frame(do.call(rbind, DEG_NEO_ADJ_Seurat))

# Distribution of log2FC
hist(DEG_NEO_ADJ_Seurat_df$avg_log2FC, breaks = 100)

DEG_NEO_ADJ_Seurat_df_sig <- DEG_NEO_ADJ_Seurat_df %>%
  filter(p_val_adj < 0.01,
         abs(avg_log2FC) > 100,
         (pct.1 > 0.50 | pct.2 > 0.50))

# Saving to a file
write.table(DEG_NEO_ADJ_Seurat_df_sig,
            "/scratch/hnatri/CART/PD1_CART_DEG_NEO_ADJ_Seurat_sig.tsv",
            sep = "\t", quote = F, row.names = F)

# Plotting
table(DEG_NEO_ADJ_df_sig$celltype) %>% as.data.frame() %>%
  ggplot(aes(x = reorder(Var1, -Freq), y = Freq, fill = Var1)) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = plot_colors) +
    theme_classic2() +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
    NoLegend() +
    xlab("Cell type") +
    ylab("# DEGs, NEO vs. ADJ, all timepoints")

table(DEG_NEO_ADJ_Seurat_df_sig$celltype) %>% as.data.frame() %>%
  ggplot(aes(x = reorder(Var1, -Freq), y = Freq, fill = Var1)) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = plot_colors) +
    theme_classic2() +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
    NoLegend() +
    xlab("Cell type") +
    ylab("# DEGs, NEO vs. ADJ, all timepoints, Seurat")

# Top DEGs for each cluster
DEG_NEO_ADJ_df_sig %>% group_by(group) %>% slice_max(order_by = auc, n = 10)
DEG_NEO_ADJ_Seurat_df_sig %>% group_by(celltype) %>% slice_max(order_by = avg_log2FC, n = 10)

# For each cluster, top DEGs between ADJ and CTRL
DEG_ADJ_CTRL <- lapply(unique(seurat_data$celltype), function(xx){
  data_subset <- subset(seurat_data, subset = celltype == xx)
  if (all((c("ADJ", "CTRL") %in% data_subset$Treatment) == c(T, T))){
    markers <- presto::wilcoxauc(data_subset,
                                 group_by = "Treatment",
                                 groups_use = c("ADJ", "CTRL"),
                                 assay = "data",
                                 seurat_assay = "RNA_human")
    #Idents(data_subset) <- data_subset$Treatment
    #markers <- FindMarkers(data_subset,
    #                       group.by = "celltype",
    #                       ident.1 = "ADJ",
    #                       ident.2 = "CTRL")
    markers$celltype <- xx
    
    return(markers)
  } else {
    return(NULL)
  }
})
names(DEG_ADJ_CTRL) <- unique(seurat_data$celltype)
DEG_ADJ_CTRL[sapply(DEG_ADJ_CTRL, is.null)] <- NULL

DEG_ADJ_CTRL_df <- as.data.frame(do.call(rbind, DEG_ADJ_CTRL))

DEG_ADJ_CTRL_df_sig <- DEG_ADJ_CTRL_df %>%
  filter(group == "NEO",
         padj < 0.01,
         auc > 0.6,
         (pct_in > 50 | pct_out > 50))

# Saving to a file
write.table(DEG_ADJ_CTRL_df_sig,
            "/scratch/hnatri/CART/PD1_CART_DEG_ADJ_CTRL_sig.tsv",
            sep = "\t", quote = F, row.names = F)

# Comparing to the Seurat function
DEG_ADJ_CTRL_Seurat <- lapply(unique(seurat_data$celltype), function(xx){
  data_subset <- subset(seurat_data, subset = celltype == xx)
  Idents(data_subset) <- data_subset$Treatment
  if (all((c("ADJ", "CTRL") %in% data_subset$Treatment) == c(T, T))){
    markers <- FindMarkers(data_subset,
                           ident.1 = "ADJ",
                           ident.2 = "CTRL",
                           assay = "RNA_human",
                           verbose = F)
    markers$feature <- rownames(markers)
    markers$celltype <- xx
    
    return(markers)
  } else {
    return(NULL)
  }
})
names(DEG_ADJ_CTRL_Seurat) <- unique(seurat_data$celltype)
DEG_ADJ_CTRL_Seurat[sapply(DEG_ADJ_CTRL_Seurat, is.null)] <- NULL

DEG_ADJ_CTRL_Seurat_df <- as.data.frame(do.call(rbind, DEG_ADJ_CTRL_Seurat))

# Distribution of log2FC
hist(DEG_ADJ_CTRL_Seurat_df$avg_log2FC, breaks = 100)

DEG_ADJ_CTRL_Seurat_df_sig <- DEG_ADJ_CTRL_Seurat_df %>%
  filter(p_val_adj < 0.01,
         abs(avg_log2FC) > 100,
         (pct.1 > 0.50 | pct.2 > 0.50))

# Saving to a file
write.table(DEG_ADJ_CTRL_Seurat_df_sig,
            "/scratch/hnatri/CART/PD1_CART_DEG_ADJ_CTRL_Seurat_sig.tsv",
            sep = "\t", quote = F, row.names = F)

# Plotting
table(DEG_ADJ_CTRL_df_sig$celltype) %>% as.data.frame() %>%
  ggplot(aes(x = reorder(Var1, -Freq), y = Freq, fill = Var1)) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = plot_colors) +
    theme_classic2() +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
    NoLegend() +
    xlab("Cell type") +
    ylab("# DEGs, ADJ vs. CTRL, all timepoints, Presto")

table(DEG_ADJ_CTRL_Seurat_df_sig$celltype) %>% as.data.frame() %>%
  ggplot(aes(x = reorder(Var1, -Freq), y = Freq, fill = Var1)) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = plot_colors) +
    theme_classic2() +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
    NoLegend() +
    xlab("Cell type") +
    ylab("# DEGs, ADJ vs. CTRL, all timepoints, Seurat")

# Top DEGs for each cluster
DEG_ADJ_CTRL_df_sig %>% group_by(group) %>% slice_max(order_by = auc, n = 10)
DEG_ADJ_CTRL_Seurat_df_sig %>% group_by(celltype) %>% slice_max(order_by = avg_log2FC, n = 10)

# Overlap of DEGs between NEO vs. CTRL and NEO vs. ADJ
intersect(DEG_NEO_ADJ_df_sig$feature, DEG_NEO_CTRL_df_sig$feature)

inner_join(DEG_NEO_ADJ_df_sig, DEG_NEO_CTRL_df_sig, by = c('feature','celltype'))

write.table(inner_join(DEG_NEO_ADJ_df_sig, DEG_NEO_CTRL_df_sig, by = c('feature','celltype')),
            "/scratch/hnatri/CART/PD1_CART_DEG_NEO_ADJ_CTRL_overlap.tsv",
            sep = "\t", quote = F, row.names = F)

# Overlap with Seurat
intersect(DEG_NEO_ADJ_Seurat_df_sig$feature, DEG_NEO_CTRL_Seurat_df_sig$feature)

inner_join(DEG_NEO_ADJ_Seurat_df_sig, DEG_NEO_CTRL_Seurat_df_sig,
           by = c('feature','celltype'))

write.table(inner_join(DEG_NEO_ADJ_Seurat_df_sig, DEG_NEO_CTRL_Seurat_df_sig,
                       by = c('feature','celltype')),
            "/scratch/hnatri/CART/PD1_CART_DEG_NEO_ADJ_CTRL_overlap.tsv",
            sep = "\t", quote = F, row.names = F)

```

### DEGs between treatment groups by timepoint

Comparing NEO vs. CTRL on Day 12 and NEO vs. CTRL, NEO vs. ADJ, and ADJ vs. CTRL
on Day 16.

```{r}

# A function for running the analysis
get_degs <- function(group1, group2, day){

  # For each cluster, top DEGs between group1 and group2 on day x
  # Using the Seurat function
  DEG_list <- lapply(unique(seurat_data$celltype), function(xx){
    data_subset <- subset(seurat_data, subset = celltype == xx)
    data_subset <- subset(data_subset, subset = Day == day)
    Idents(data_subset) <- data_subset$Treatment
    if (all((c(group1, group2) %in% data_subset$Treatment) == c(T, T))){
      markers <- FindMarkers(data_subset,
                             ident.1 = group1,
                             ident.2 = group2,
                             assay = "RNA_human",
                             verbose = F)
      markers$feature <- rownames(markers)
      markers$celltype <- xx
      
      return(markers)
    } else {
      return(NULL)
    }
  })
  names(DEG_list) <- unique(seurat_data$celltype)
  DEG_list[sapply(DEG_list, is.null)] <- NULL
  
  DEG_df <- as.data.frame(do.call(rbind, DEG_list))
  
  DEG_df_sig <- DEG_df %>%
    filter(p_val_adj < 0.01,
           abs(avg_log2FC) > 100,
           (pct.1 > 0.50 | pct.2 > 0.50))
  
  # Saving to a file
  filename <- paste0("/scratch/hnatri/CART/PD1_CART_DEG_", group1, "_", group2,
                     "_", day, "_Seurat_sig.tsv")
  write.table(DEG_df_sig, filename,
              sep = "\t", quote = F, row.names = F)
  
  return(DEG_df_sig)
}

# For each comparison and timepoint, calling DEGs and plotting
DEG_NEO_CTRL_D12 <- get_degs(group1 = "NEO",
                             group2 = "CTRL", 
                             day = 12)

DEG_NEO_CTRL_D16 <- get_degs(group1 = "NEO",
                             group2 = "CTRL", 
                             day = 16)

DEG_NEO_ADJ_D16 <- get_degs(group1 = "NEO",
                            group2 = "ADJ", 
                            day = 16)

DEG_ADJ_CTRL_D16 <- get_degs(group1 = "ADJ",
                             group2 = "CTRL", 
                             day = 16)

length(unique(DEG_NEO_CTRL_D12$feature))
length(unique(DEG_NEO_CTRL_D16$feature))
length(unique(DEG_NEO_ADJ_D16$feature))
length(unique(DEG_ADJ_CTRL_D16$feature))

# Plotting
table(DEG_NEO_CTRL_D12$celltype) %>% as.data.frame() %>%
  ggplot(aes(x = reorder(Var1, -Freq), y = Freq, fill = Var1)) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = plot_colors) +
    theme_classic2() +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
    NoLegend() +
    xlab("Cell type") +
    ylab("# DEGs NEO vs. CTRL, Day 12")

table(DEG_NEO_CTRL_D16$celltype) %>% as.data.frame() %>%
  ggplot(aes(x = reorder(Var1, -Freq), y = Freq, fill = Var1)) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = plot_colors) +
    theme_classic2() +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
    NoLegend() +
    xlab("Cell type") +
    ylab("# DEGs NEO vs. CTRL, Day 16")

table(DEG_NEO_ADJ_D16$celltype) %>% as.data.frame() %>%
  ggplot(aes(x = reorder(Var1, -Freq), y = Freq, fill = Var1)) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = plot_colors) +
    theme_classic2() +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
    NoLegend() +
    xlab("Cell type") +
    ylab("# DEGs NEO vs. ADJ, Day 16")

table(DEG_ADJ_CTRL_D16$celltype) %>% as.data.frame() %>%
  ggplot(aes(x = reorder(Var1, -Freq), y = Freq, fill = Var1)) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = plot_colors) +
    theme_classic2() +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
    NoLegend() +
    xlab("Cell type") +
    ylab("# DEGs ADJ vs. CTRL, Day 12=6")

# Top DEGs for each cluster
#DEG_df_sig %>% group_by(celltype) %>% slice_max(order_by = avg_log2FC, n = 10)
```

### Looking at DEGs unique to/shared between groups

```{r}
# Overlap of significant DEGs in each comparison
input_list <- list("NEO vs. CTRL, Day 12" = DEG_NEO_CTRL_D12$feature,
                   "NEO vs. CTRL, Day 16" = DEG_NEO_CTRL_D16$feature,
                   "NEO vs. ADJ, Day 16" = DEG_NEO_ADJ_D16$feature,
                   "ADJ vs. CTRL, Day 16" = DEG_ADJ_CTRL_D16$feature)

upset(fromList(input_list), order.by = "freq")

# Plotting overlaps without RP and MT genes
RBMTgenes <- grep(pattern = "^RP[SL]|^MRP[SL]|^MT-", x = rownames(seurat_data@assays$RNA_human@data), value = TRUE)
    
input_list <- lapply(input_list, function(xx){
  setdiff(xx, RBMTgenes)
})

upset(fromList(input_list), order.by = "freq")

# Looking at genes in intersections
# Unique to ADJ_CTRL_D16
setdiff(DEG_ADJ_CTRL_D16$feature, c(DEG_NEO_ADJ_D16$feature,
                                    DEG_NEO_CTRL_D16$feature,
                                    DEG_NEO_CTRL_D12$feature,
                                    RBMTgenes))
# Unique to DEG_NEO_CTRL_D16
setdiff(DEG_NEO_CTRL_D16$feature, c(DEG_NEO_ADJ_D16$feature,
                                    DEG_ADJ_CTRL_D16$feature,
                                    DEG_NEO_CTRL_D12$feature,
                                    RBMTgenes))
# Unique to DEG_NEO_ADJ_D16
setdiff(DEG_NEO_ADJ_D16$feature, c(DEG_NEO_CTRL_D16$feature,
                                   DEG_ADJ_CTRL_D16$feature,
                                   DEG_NEO_CTRL_D12$feature,
                                   RBMTgenes))

# Unique to DEG_NEO_ADJ_D16
setdiff(DEG_NEO_CTRL_D12$feature, c(DEG_NEO_CTRL_D16$feature,
                                    DEG_ADJ_CTRL_D16$feature,
                                    DEG_NEO_ADJ_D16$feature,
                                    RBMTgenes))

# Shared between NEO vs. ADJ and ADJ vs. CTRL, without RB/MT genes
setdiff(intersect(DEG_NEO_ADJ_D16$feature, DEG_ADJ_CTRL_D16$feature), RBMTgenes)

```

### Plotting SPP1

```{r, fig.height = 12}
# Looking at SPP1
DEG_NEO_ADJ_D16 %>% filter(feature == "SPP1")
VlnPlot(seurat_data,
        features = "SPP1",
        group.by = "celltype",
        split.by = "Treatment_Day",
        assay = "RNA_human",
        layer = "data",
        pt.size = 0,
        log = T) +
  ylim(0, 1000) +
  coord_flip()

```


### Human T cell canonical markers

```{r}

# Canonical T cell markers
gs4_deauth()
canonical_markers  <- gs4_get("https://docs.google.com/spreadsheets/d/186PIcU3Jb0Xm3IuZgENEChluykvwsK4JnHF8FbeJASA/edit?usp=sharing")
sheet_names(canonical_markers)
canonical_markers <- read_sheet(canonical_markers, sheet = "T cells, gene sets")
head(canonical_markers)

# Overlapping with cluster markers
sig_markers %>% filter(feature %in% canonical_markers$RNA)
canonical_markers %>% filter(RNA %in% sig_markers$feature) %>% as.data.frame()

# Overlapping with DEGs
DEG_NEO_CTRL_df_sig %>% filter(feature %in% canonical_markers$RNA)
DEG_NEO_ADJ_df_sig %>% filter(feature %in% canonical_markers$RNA)

# Overlapping with DEGs from the Seurat analysis
DEG_NEO_CTRL_Seurat_df_sig %>% filter(feature %in% canonical_markers$RNA)
DEG_NEO_ADJ_Seurat_df_sig %>% filter(feature %in% canonical_markers$RNA)

```
