---
title: "GSEA pathway analysis on the Kluc tumor CD45+ scRNAseq data"
author: "heinin"
date: "2024-03-22"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

### Introduction

Running GSEA for each cell and comparing between groups.

### Packages and environment variables

```{r, warning=F, message=F}

suppressPackageStartupMessages({
  #library(cli)
  library(Seurat)
  library(SeuratObject)
  library(SeuratDisk)
  library(tidyverse)
  library(tibble)
  library(ggplot2)
  library(ggpubr)
  library(ggrepel)
  library(workflowr)
  library(dittoSeq)
  library(googlesheets4)
  library(escape)
  library(pheatmap)})

setwd("/home/hnatri/PD1_mm/")
set.seed(9999)
options(ggrepel.max.overlaps = Inf)

# Colors, themes, cell type markers, and plot functions
source("/home/hnatri/PD1_mm/code/utilities.R")
source("/home/hnatri/PD1_mm/code/PD1_mm_themes.R")
source("/home/hnatri/PD1_mm/code/CART_plot_functions.R")

```

### Importing data

```{r}

#seurat_data <- readRDS("/tgen_labs/banovich/BCTCSF/PD1_mm_Seurat/PD1_mm_Seurat_merged.Rds")
#seurat_data$Treatment_Day <- paste0(seurat_data$Treatment, "_", seurat_data$Day)

```

### Running GSEA using R/escape

GSEA was run using code/GSEA.R.

### Importing GSEA outputs

```{r}

seurat_data <- readRDS("/tgen_labs/banovich/BCTCSF/PD1_mm_Seurat/PD1_mm_Seurat_GSEA.Rds")
res <- readRDS("/scratch/hnatri/CART/KlucCD45pos_GSEA_res.rds")

```

### Significance testing

```{r}

res <- seurat_data@meta.data %>%
  dplyr::select(c("orig.ident", "Treatment_Day", colnames(res)))

sig_res <- data.frame(getSignificance(res,
                                      group = "Treatment_Day",
                                      fit = "ANOVA"))
sig_res$pathways <- rownames(sig_res)
sig_res <- sig_res %>% filter(FDR < 0.05) %>% arrange(FDR)

dim(sig_res)

sig_res %>% dplyr::select(c("pathways"))

```

### Plotting results

```{r, fig.width=14, fig.height=8}

DefaultAssay(seurat_data) <- "RNA_human"
seurat_data$Treatment_Day <- factor(seurat_data$Treatment_Day,
                                    levels = c("NEO_12", "NEO_16",
                                               "CTRL_12", "CTRL_16",
                                               "ADJ_16"))

dittoHeatmap(seurat_data,
             genes = NULL,
             metas = sig_res$pathways, 
             heatmap.colors = colorRampPalette(c("dodgerblue4", "white", "tomato"))(100),
             annot.colors = treatment_day_col,
             annot.by = "Treatment_Day",
             order.by = "Treatment_Day",
             cluster_cols = F,
             fontsize = 7)

dittoHeatmap(seurat_data,
             genes = NULL,
             metas = sig_res$pathways, 
             heatmap.colors = colorRampPalette(c("dodgerblue4", "white", "tomato"))(100),
             annot.colors = treatment_day_col,
             annot.by = "Treatment_Day",
             cluster_cols = T,
             fontsize = 7)

```

```{r, fig.width=16, fig.height=28}
VlnPlot(seurat_data,
        features = sig_res$pathways,
        group.by = "celltype",
        split.by = "Treatment_Day",
        cols = treatment_day_col,
        ncol = 1,
        pt.size = 0) &
  theme_classic() &
  xlab("") &
  NoLegend()
```

### Plotting by celltype and treatment only

```{r, fig.width=28, fig.height=42}
VlnPlot(seurat_data,
        features = sig_res$pathways,
        group.by = "celltype",
        split.by = "Treatment",
        cols = treatment_day_col,
        ncol = 1,
        pt.size = 0) &
  theme_classic() &
  xlab("") &
  NoLegend()
```

### Plotting by celltype and timepoint only

```{r, fig.width=28, fig.height=42}
VlnPlot(seurat_data,
        features = sig_res$pathways,
        group.by = "celltype",
        split.by = "Day",
        cols = c("deeppink3", "aquamarine3"),
        ncol = 1,
        pt.size = 0) &
  theme_classic() &
  xlab("") &
  NoLegend()
```

### Plotting by treatment only

```{r, fig.width=3, fig.height=24}
VlnPlot(seurat_data,
        features = sig_res$pathways,
        group.by = "Treatment",
        cols = treatment_col,
        ncol = 1,
        pt.size = 0) &
  theme_classic() &
  xlab("") &
  NoLegend()
```

### Plotting by timepoint only

```{r, fig.width=3, fig.height=24}
VlnPlot(seurat_data,
        features = sig_res$pathways,
        group.by = "Day",
        cols = c("deeppink3", "aquamarine3"),
        ncol = 1,
        pt.size = 0) &
  theme_classic() &
  xlab("") &
  NoLegend()
```

### Testing with myeloid cell types only

```{r, fig.height=10, fig.width=12, warning=F, message=F}

seurat_data_myeloid <- subset(seurat_data, subset = celltype %in% paste0("M", seq(1, 9)))

res_myeloid <- seurat_data_myeloid@meta.data %>%
  dplyr::select(c("orig.ident", "Treatment_Day", colnames(res)))

sig_res_myeloid <- data.frame(getSignificance(res_myeloid,
                                              group = "Treatment_Day",
                                              fit = "ANOVA"))
sig_res_myeloid$pathways <- rownames(sig_res_myeloid)
sig_res_myeloid <- sig_res_myeloid %>% filter(FDR < 0.05) %>% arrange(FDR)

dim(sig_res_myeloid)

sig_res_myeloid %>% dplyr::select(c("pathways"))

write.table(sig_res_myeloid, "/home/hnatri/PD1_mm/GSEA_myeloid_sig.tsv",
            sep = "\t", quote = F, row.names = F)

# Overlap of significant pathways for the whole data vs. myeloid only
intersect(sig_res_myeloid$pathways, sig_res$pathways)

# Checking significance by Treatment, Day and celltype in myeloid cells
seurat_data_myeloid$Treatment_Day_celltype <- paste0(seurat_data_myeloid$Treatment_Day,
                                                     "_", seurat_data_myeloid$celltype)
res_myeloid_celltype <- seurat_data_myeloid@meta.data %>%
  dplyr::select(c("orig.ident", "Treatment_Day_celltype", colnames(res)))

sig_res_myeloid_celltype <- data.frame(getSignificance(res_myeloid_celltype,
                                              group = "Treatment_Day_celltype",
                                              fit = "ANOVA"))
sig_res_myeloid_celltype$pathways <- rownames(sig_res_myeloid_celltype)
sig_res_myeloid_celltype <- sig_res_myeloid_celltype %>%
  filter(FDR < 0.05) %>% arrange(FDR)

unique(sig_res_myeloid_celltype$pathways)

# Only keeping comparisons for the same cell type
sig_res_myeloid_celltype <- t(sig_res_myeloid_celltype)
sig_res_myeloid_celltype <- sig_res_myeloid_celltype %>% as.data.frame() %>%
  rownames_to_column(var = "comparison") %>%
  separate(col = comparison, sep = "v", into = c("group1", "group2")) %>%
  separate(col = group1, sep = "_", into = c("group1", "day1", "celltype1")) %>%
  separate(col = group2, sep = "_", into = c("group2", "day2", "celltype2")) %>%
  separate(col = celltype2, sep = "\\.", into = c("celltype2", "valtype")) %>% 
  dplyr::filter(celltype1 == celltype2 | group1 == "f." | group1 == "p." | group1 == "pathways")

dim(sig_res_myeloid_celltype)

write.table(sig_res_myeloid_celltype, "/home/hnatri/PD1_mm/GSEA_myeloid_celltype_sig.tsv",
            sep = "\t", quote = F, row.names = F)

# Heatmap of p-values
sig_res_myeloid_celltype <- sig_res_myeloid_celltype %>% filter(!group1 %in% c("f.", "p.")) %>%
  select(!c("valtype"))

colnames(sig_res_myeloid_celltype) <- c(colnames(sig_res_myeloid_celltype)[1:6], as.character(sig_res_myeloid_celltype[nrow(sig_res_myeloid_celltype),][7:ncol(sig_res_myeloid_celltype)]))

# Bottom, left, top, right
#par(mar=c(2,12,2,3)) 

sig_res_myeloid_celltype %>% unite("comparison", group1:celltype2, sep = "_") %>% 
  column_to_rownames(var = "comparison") %>%
  filter(row_number() <= n()-1) %>%
  mutate_if(is.character, as.numeric) %>% 
  filter_all(any_vars(is.numeric(.) & . < 0.05)) %>%
  as.matrix() %>%
  pheatmap(color = colorRampPalette(brewer.pal(n = 7, name = "RdYlBu"))(100),
           fontsize = 7,
           angle_col = "45",
           display_numbers = T,
           number_format = "%.2f", number_color = "black")

```

### Plotting results for myeloid

```{r, fig.width=14, fig.height=6}

dittoHeatmap(seurat_data_myeloid,
             genes = NULL,
             metas = sig_res_myeloid$pathways, 
             heatmap.colors = colorRampPalette(c("dodgerblue4", "white", "tomato"))(100),
             annot.colors = treatment_day_col,
             annot.by = "Treatment_Day",
             order.by = "Treatment_Day",
             cluster_cols = F,
             fontsize = 7)

dittoHeatmap(seurat_data_myeloid,
             genes = NULL,
             metas = sig_res_myeloid$pathways, 
             heatmap.colors = colorRampPalette(c("dodgerblue4", "white", "tomato"))(100),
             annot.colors = treatment_day_col,
             annot.by = "Treatment_Day",
             cluster_cols = T,
             fontsize = 7)

```

```{r, fig.width=16, fig.height=28}
VlnPlot(seurat_data_myeloid,
        features = sig_res$pathways,
        group.by = "celltype",
        split.by = "Treatment_Day",
        cols = treatment_day_col,
        ncol = 1,
        pt.size = 0) &
  theme_classic() &
  xlab("") &
  NoLegend()
```

### Plotting by celltype and treatment (myeloid)

```{r, fig.width=28, fig.height=42}
VlnPlot(seurat_data_myeloid,
        features = sig_res$pathways,
        group.by = "celltype",
        split.by = "Treatment",
        cols = treatment_day_col,
        ncol = 1,
        pt.size = 0) &
  theme_classic() &
  xlab("") &
  NoLegend()
```

### Plotting by celltype and timepoint only (myeloid)

```{r, fig.width=28, fig.height=42}
VlnPlot(seurat_data_myeloid,
        features = sig_res$pathways,
        group.by = "celltype",
        split.by = "Day",
        cols = c("deeppink3", "aquamarine3"),
        ncol = 1,
        pt.size = 0) &
  theme_classic() &
  xlab("") &
  NoLegend()
```

### Plotting by treatment only (myeloid)

```{r, fig.width=3, fig.height=24}
VlnPlot(seurat_data_myeloid,
        features = sig_res$pathways,
        group.by = "Treatment",
        cols = treatment_col,
        ncol = 1,
        pt.size = 0) &
  theme_classic() &
  xlab("") &
  NoLegend()
```

### Plotting by timepoint only (myeloid)

```{r, fig.width=3, fig.height=24}
VlnPlot(seurat_data_myeloid,
        features = sig_res$pathways,
        group.by = "Day",
        cols = c("deeppink3", "aquamarine3"),
        ncol = 1,
        pt.size = 0) &
  theme_classic() &
  xlab("") &
  NoLegend()
```
