---
title: "GSEA pathway analysis on the Kluc tumor CD45+ scRNAseq data"
author: "heinin"
date: "2024-03-22"
output:
  workflowr::wflow_html:
    code_folding: hide
  html_notebook: default
editor_options:
  chunk_output_type: console
---

## Introduction

Running GSEA for each cell and comparing between groups.

## Packages and environment variables

```{r, warning=F, message=F}

suppressPackageStartupMessages({
  library(Seurat)
  library(SeuratObject)
  library(SeuratDisk)
  library(tidyverse)
  library(tibble)
  library(ggplot2)
  library(ggpubr)
  library(ggrepel)
  library(workflowr)
  library(dittoSeq)
  library(googlesheets4)
  library(escape)
  library(pheatmap)})

setwd("/home/hnatri/PD1_mm/")
set.seed(9999)
options(ggrepel.max.overlaps = Inf)

# Colors, themes, cell type markers, and plot functions
source("/home/hnatri/PD1_mm/code/utilities.R")
source("/home/hnatri/PD1_mm/code/PD1_mm_themes.R")
source("/home/hnatri/PD1_mm/code/CART_plot_functions.R")

```

## Importing data

```{r, warning=F, message=F, eval=F}

#seurat_data <- readRDS("/tgen_labs/banovich/BCTCSF/PD1_mm_Seurat/PD1_mm_Seurat_merged.Rds")
#seurat_data$Treatment_Day <- paste0(seurat_data$Treatment, "_", seurat_data$Day)

```

## Running GSEA using R/escape

GSEA was run using code/GSEA.R.

## Importing GSEA outputs

```{r, warning=F, message=F}

seurat_data <- readRDS("/tgen_labs/banovich/BCTCSF/PD1_mm_Seurat/PD1_mm_Seurat_GSEA.Rds")

seurat_data <- subset(seurat_data, subset = Day == 16)

sort(as.character(unique(seurat_data$celltype)))

seurat_data$Treatment_Day <- factor(seurat_data$Treatment_Day, levels = c("CTRL_16", "NEO_16", "ADJ_16"))
Idents(seurat_data) <- seurat_data$Treatment_Day

```

## Differentially activated pathways between ADJ and NEO, all cell types, Day 16

```{r, message = F, warning = F, results = 'hide', fig.width = 8, fig.height = 8}

gsea_res <- FindMarkers(seurat_data,
                        assay = "escapeGSVA_normalized",
                        ident.1 = "NEO_16",
                        ident.2 = "ADJ_16",
                        min.pct = 0,
                        logfc.threshold = 0)

gsea_res$pathway <- rownames(gsea_res)

#head(gsea_res)
#hist(gsea_res$avg_log2FC)
#hist(gsea_res$p_val_adj)

res_sig <- gsea_res %>% filter(abs(avg_log2FC) > 0.1,
                               p_val_adj < 0.05)


res_sig %>%
  #arrange(c("p_val_adj", avg_log2FC)) %>%
  #filter(dense_rank(avg_log2FC) <= 10 | dense_rank(desc(avg_log2FC)) <= 10) %>%
  #filter(abs(avg_log2FC) > 0.1) %>%
  mutate(sign = sign(avg_log2FC),
         signstr = if_else(sign == 1, "ADJ_16", "NEO_16")) %>%
  ggplot(aes(x = avg_log2FC, y = reorder(pathway, avg_log2FC), fill = signstr)) +
    geom_bar(stat = "identity") +
    #geom_col(width = 0.85) +
    scale_fill_manual(values = c("deeppink3", "aquamarine3")) +
    theme_classic() +
    #manuscript_theme +
    ylab("") +
    xlab(expression(Delta ~ " average log2(FC)")) +
    ggtitle("ADJ Day 16 vs. NEO Day 16")


```

## Differentially activated pathways between NEO and CTRL, all cell types, Day 16

```{r, message = F, warning = F, results = 'hide', fig.width=16, fig.height=18}

gsea_res <- FindMarkers(seurat_data,
                        assay = "escapeGSVA_normalized", 
                        ident.1 = "NEO_16",
                        ident.2 = "CTRL_16",
                        min.pct = 0,
                        logfc.threshold = 0)

gsea_res$pathway <- rownames(gsea_res)

res_sig <- gsea_res %>% filter(abs(avg_log2FC) > 0.1,
                               p_val_adj < 0.05)

res_sig %>%
  #filter(abs(avg_log2FC) > 0.15) %>%
  filter(dense_rank(avg_log2FC) <= 10 | dense_rank(desc(avg_log2FC)) <= 10) %>%
  mutate(sign = sign(avg_log2FC),
         signstr = if_else(sign == 1, "NEO_16", "CTRL_16")) %>%
  ggplot(aes(x = avg_log2FC, y = reorder(pathway, avg_log2FC), fill = signstr)) +
    geom_bar(stat = "identity") +
    #geom_col(width = 0.85) +
    scale_fill_manual(values = c("deeppink3", "aquamarine3")) +
    theme_classic() +
    #manuscript_theme +
    ylab("") +
    xlab(expression(Delta ~ " average log2(FC)")) +
    ggtitle("NEO Day 16 vs. CTRL Day 16") +
    theme(axis.text = element_text(size = 12))

```

## Differentially activated pathways between ADJ and CTRL, all cell types, Day 16

```{r, message = F, warning = F, results = 'hide', fig.width=16, fig.height=18}

gsea_res <- FindMarkers(seurat_data,
                        assay = "escapeGSVA_normalized", 
                        ident.1 = "ADJ_16",
                        ident.2 = "CTRL_16",
                        min.pct = 0,
                        logfc.threshold = 0)

gsea_res$pathway <- rownames(gsea_res)

res_sig <- gsea_res %>% filter(abs(avg_log2FC) > 0.1,
                               p_val_adj < 0.05)

res_sig %>%
  #filter(abs(avg_log2FC) > 0.15) %>%
  filter(dense_rank(avg_log2FC) <= 10 | dense_rank(desc(avg_log2FC)) <= 10) %>%
  mutate(sign = sign(avg_log2FC),
         signstr = if_else(sign == 1, "ADJ_16", "CTRL_16")) %>%
  ggplot(aes(x = avg_log2FC, y = reorder(pathway, avg_log2FC), fill = signstr)) +
    geom_bar(stat = "identity") +
    #geom_col(width = 0.85) +
    scale_fill_manual(values = c("deeppink3", "aquamarine3")) +
    theme_classic() +
    #manuscript_theme +
    ylab("") +
    xlab(expression(Delta ~ " average log2(FC)")) +
    ggtitle("ADJ Day 16 vs. CTRL Day 16") +
    theme(axis.text = element_text(size = 12))

```

## Differentially activated pathways between NEO and CTRL, all cell types, Day 12

```{r, message = F, warning = F, results = 'hide', fig.width=16, fig.height=18, eval = F}

gsea_res <- FindMarkers(seurat_data,
                        assay = "escapeGSVA_normalized", 
                        ident.1 = "NEO_12",
                        ident.2 = "CTRL_12",
                        min.pct = 0,
                        logfc.threshold = 0)

gsea_res$pathway <- rownames(gsea_res)

res_sig <- gsea_res %>% filter(abs(avg_log2FC) > 0.1,
                               p_val_adj < 0.05)

res_sig %>%
  #filter(abs(avg_log2FC) > 0.15) %>%
  filter(dense_rank(avg_log2FC) <= 10 | dense_rank(desc(avg_log2FC)) <= 10) %>%
  #filter(dense_rank(abs(avg_log2FC)) <= 20) %>%
  mutate(sign = sign(avg_log2FC),
         signstr = if_else(sign == 1, "NEO_12", "CTRL_12")) %>%
  ggplot(aes(x = avg_log2FC, y = reorder(pathway, avg_log2FC), fill = signstr)) +
    geom_bar(stat = "identity") +
    #geom_col(width = 0.85) +
    scale_fill_manual(values = c("deeppink3", "aquamarine3")) +
    theme_classic() +
    #manuscript_theme +
    ylab("") +
    xlab(expression(Delta ~ " average log2(FC)")) +
    ggtitle("NEO Day 12 vs. CTRL Day 12") +
    theme(axis.text = element_text(size = 12))

```

## Significance testing across myeloid celltypes (M1-7, Neut)

### NEO vs. ADJ, Day 16

```{r, message = F, warning = F, results = 'hide', fig.width = 8, fig.height = 4}

unique(seurat_data$celltype)
myeloid <- subset(seurat_data, subset = celltype %in% c(grep("^M", unique(seurat_data$celltype), value = T), c("Neut1")))

gsea_res <- FindMarkers(myeloid,
                        assay = "escapeGSVA_normalized", 
                        ident.1 = "NEO_16",
                        ident.2 = "ADJ_16",
                        min.pct = 0,
                        logfc.threshold = 0)

gsea_res$pathway <- rownames(gsea_res)

res_sig <- gsea_res %>% filter(abs(avg_log2FC) > 0.1,
                               p_val_adj < 0.05)

res_sig %>%
  #filter(abs(avg_log2FC) > 0.1) %>%
  filter(dense_rank(avg_log2FC) <= 10 | dense_rank(desc(avg_log2FC)) <= 10) %>%
  mutate(sign = sign(avg_log2FC),
         signstr = if_else(sign == 1, "NEO_16", "ADJ_16")) %>%
  ggplot(aes(x = avg_log2FC, y = reorder(pathway, avg_log2FC), fill = signstr)) +
    geom_bar(stat = "identity") +
    #geom_col(width = 0.85) +
    scale_fill_manual(values = c("deeppink3", "aquamarine3")) +
    theme_classic() +
    #manuscript_theme +
    ylab("") +
    xlab(expression(Delta ~ " average log2(FC)")) +
    ggtitle("NEO Day 16 vs. ADJ Day 16, myeloid") +
    theme(axis.text = element_text(size = 12))

```

### NEO vs. CTRL, Day 16

```{r, message = F, warning = F, results = 'hide', fig.width=14, fig.height=14}

gsea_res <- FindMarkers(myeloid,
                        assay = "escapeGSVA_normalized", 
                        ident.1 = "NEO_16",
                        ident.2 = "CTRL_16",
                        min.pct = 0,
                        logfc.threshold = 0)

gsea_res$pathway <- rownames(gsea_res)

res_sig <- gsea_res %>% filter(abs(avg_log2FC) > 0.1,
                               p_val_adj < 0.05)

res_sig %>%
  #filter(abs(avg_log2FC) > 0.15) %>%
  filter(dense_rank(rev(avg_log2FC)) <= 20) %>% #| dense_rank(desc(avg_log2FC)) <= 10) %>%
  mutate(sign = sign(avg_log2FC),
         signstr = if_else(sign == 1, "NEO_16", "CTRL_16")) %>%
  ggplot(aes(x = avg_log2FC, y = reorder(pathway, avg_log2FC), fill = signstr)) +
    geom_bar(stat = "identity") +
    #geom_col(width = 0.85) +
    scale_fill_manual(values = c("deeppink3", "aquamarine3")) +
    theme_classic() +
    #manuscript_theme +
    ylab("") +
    xlab(expression(Delta ~ " average log2(FC)")) +
    ggtitle("NEO Day 16 vs. CTRL Day 16, myeloid") +
    theme(axis.text = element_text(size = 12))

```

### ADJ vs. CTRL, Day 16

```{r, message = F, warning = F, fig.width=14, fig.height=14}

gsea_res2 <- FindMarkers(myeloid,
                        assay = "escapeGSVA_normalized", 
                        ident.1 = "ADJ_16",
                        ident.2 = "CTRL_16",
                        min.pct = 0,
                        logfc.threshold = 0)

gsea_res2$pathway <- rownames(gsea_res2)

res_sig2 <- gsea_res2 %>% filter(abs(avg_log2FC) > 0.1,
                               p_val_adj < 0.05)

res_sig2 %>%
  #filter(abs(avg_log2FC) > 0.15) %>%
  filter(dense_rank(desc(avg_log2FC)) <= 20) %>%
  mutate(sign = sign(avg_log2FC),
         signstr = if_else(sign == 1, "ADJ_16", "CTRL_16")) %>%
  ggplot(aes(x = avg_log2FC, y = reorder(pathway, avg_log2FC), fill = signstr)) +
    geom_bar(stat = "identity") +
    #geom_col(width = 0.85) +
    scale_fill_manual(values = c("deeppink3", "aquamarine3")) +
    theme_classic() +
    #manuscript_theme +
    ylab("") +
    xlab(expression(Delta ~ " average log2(FC)")) +
    ggtitle("ADJ Day 16 vs. CTRL Day 16, myeloid") +
    theme(axis.text = element_text(size = 12))

length(intersect(res_sig2$pathway, res_sig$pathway))
length(setdiff(res_sig2$pathway, res_sig$pathway))
length(setdiff(res_sig$pathway, res_sig2$pathway))

```

### NEO vs. CTRL, Day 12

```{r, message = F, warning = F, results = 'hide', fig.width=15, fig.height=15, eval = F}

gsea_res <- FindMarkers(myeloid,
                        assay = "escapeGSVA_normalized", 
                        ident.1 = "NEO_12",
                        ident.2 = "CTRL_12",
                        min.pct = 0,
                        logfc.threshold = 0)

gsea_res$pathway <- rownames(gsea_res)

res_sig <- gsea_res %>% filter(abs(avg_log2FC) > 0.1,
                               p_val_adj < 0.05)

res_sig %>%
  #filter(abs(avg_log2FC) > 0.15) %>%
  filter(dense_rank(avg_log2FC) <= 16 | dense_rank(desc(avg_log2FC)) <= 4) %>%
  mutate(sign = sign(avg_log2FC),
         signstr = if_else(sign == 1, "NEO_12", "CTRL_12")) %>%
  ggplot(aes(x = avg_log2FC, y = reorder(pathway, avg_log2FC), fill = signstr)) +
    geom_bar(stat = "identity") +
    #geom_col(width = 0.85) +
    scale_fill_manual(values = c("deeppink3", "aquamarine3")) +
    theme_classic() +
    #manuscript_theme +
    ylab("") +
    xlab(expression(Delta ~ " average log2(FC)")) +
    ggtitle("NEO Day 12 vs. CTRL Day 12, myeloid") +
    theme(axis.text = element_text(size = 2))

```

## Significance testing across lymphoid celltypes (L1-6, Treg, NK)

### NEO vs. ADJ, Day 16

```{r, message = F, warning = F, results = 'hide', fig.width = 8, fig.height = 4}

unique(seurat_data$celltype)
lymphoid <- subset(seurat_data, subset = celltype %in% c(grep("^L", unique(seurat_data$celltype), value = T), c("NK", "Treg")))

gsea_res <- FindMarkers(lymphoid,
                        assay = "escapeGSVA_normalized", 
                        ident.1 = "NEO_16",
                        ident.2 = "ADJ_16",
                        min.pct = 0,
                        logfc.threshold = 0)

gsea_res$pathway <- rownames(gsea_res)

res_sig <- gsea_res %>% filter(abs(avg_log2FC) > 0.1,
                               p_val_adj < 0.05)

res_sig %>%
  #filter(abs(avg_log2FC) > 0.15) %>%
  filter(dense_rank(desc(avg_log2FC)) <= 20) %>%
  mutate(sign = sign(avg_log2FC),
         signstr = if_else(sign == 1, "NEO_16", "ADJ_16")) %>%
  ggplot(aes(x = avg_log2FC, y = reorder(pathway, avg_log2FC), fill = signstr)) +
    geom_bar(stat = "identity") +
    #geom_col(width = 0.85) +
    scale_fill_manual(values = c("deeppink3", "aquamarine3")) +
    theme_classic() +
    #manuscript_theme +
    ylab("") +
    xlab(expression(Delta ~ " average log2(FC)")) +
    ggtitle("NEO Day 16 vs. ADJ Day 16, lymphoid") +
    theme(axis.text = element_text(size = 12))

```

### NEO vs. CTRL, Day 16

```{r, message = F, warning = F, results = 'hide', fig.width = 8, fig.height = 8}

gsea_res <- FindMarkers(lymphoid,
                        assay = "escapeGSVA_normalized", 
                        ident.1 = "NEO_16",
                        ident.2 = "CTRL_16",
                        min.pct = 0,
                        logfc.threshold = 0)

gsea_res$pathway <- rownames(gsea_res)

res_sig <- gsea_res %>% filter(abs(avg_log2FC) > 0.1,
                               p_val_adj < 0.05)

res_sig %>%
  #filter(abs(avg_log2FC) > 0.15) %>%
  filter(dense_rank(desc(avg_log2FC)) <= 20) %>%
  mutate(sign = sign(avg_log2FC),
         signstr = if_else(sign == 1, "NEO_16", "CTRL_16")) %>%
  ggplot(aes(x = avg_log2FC, y = reorder(pathway, avg_log2FC), fill = signstr)) +
    geom_bar(stat = "identity") +
    #geom_col(width = 0.85) +
    scale_fill_manual(values = c("deeppink3", "aquamarine3")) +
    theme_classic() +
    #manuscript_theme +
    ylab("") +
    xlab(expression(Delta ~ " average log2(FC)")) +
    ggtitle("NEO Day 16 vs. CTRL Day 16, lymphoid") +
    theme(axis.text = element_text(size = 12))


```

### ADJ vs. CTRL, Day 16

```{r, message = F, warning = F, results = 'hide', fig.width = 8, fig.height = 8}

gsea_res <- FindMarkers(lymphoid,
                        assay = "escapeGSVA_normalized", 
                        ident.1 = "ADJ_16",
                        ident.2 = "CTRL_16",
                        min.pct = 0,
                        logfc.threshold = 0)

gsea_res$pathway <- rownames(gsea_res)

res_sig <- gsea_res %>% filter(abs(avg_log2FC) > 0.1,
                               p_val_adj < 0.05)

res_sig %>%
  #filter(abs(avg_log2FC) > 0.15) %>%
  filter(dense_rank(avg_log2FC) <= 10 | dense_rank(desc(avg_log2FC)) <= 10) %>%
  mutate(sign = sign(avg_log2FC),
         signstr = if_else(sign == 1, "ADJ_16", "CTRL_16")) %>%
  ggplot(aes(x = avg_log2FC, y = reorder(pathway, avg_log2FC), fill = signstr)) +
    geom_bar(stat = "identity") +
    #geom_col(width = 0.85) +
    scale_fill_manual(values = c("deeppink3", "aquamarine3")) +
    theme_classic() +
    #manuscript_theme +
    ylab("") +
    xlab(expression(Delta ~ " average log2(FC)")) +
    ggtitle("NEO Day 16 vs. CTRL Day 16, lymphoid") +
    theme(axis.text = element_text(size = 12))

```

## Heatmaps

### Myeloid

```{r, message = F, warning = F, results = 'hide', fig.width = 15, fig.height = 10}

myeloid_res <- FindAllMarkers(myeloid,
                              assay = "escapeGSVA_normalized", 
                              return.thresh = 0.01,
                              logfc.threshold = 0.5,
                              min.pct = 0.20,
                              only.pos = T,
                              verbose = F)

myeloid_res$pathway <- rownames(myeloid_res)

myeloid_res_sig <- myeloid_res %>% filter(abs(avg_log2FC) > 0.1,
                               p_val_adj < 0.05)

plot_features <- myeloid_res_sig %>%  group_by(cluster) %>% slice_max(order_by = abs(avg_log2FC), n = 20)

# Subset for heatmap
myeloid_subset <- subset(myeloid, downsample = 30000/length(unique(myeloid$celltype)))

myeloid_subset <- ScaleData(myeloid_subset,
                            assay = "escapeGSVA_normalized")

DoHeatmap(
  myeloid_subset,
  features = unique(plot_features$gene),
  group.by = "Treatment_Day",
  group.bar = TRUE,
  assay = "escapeGSVA_normalized")

```

### Lymphoid

```{r, message = F, warning = F, results = 'hide', fig.width = 15, fig.height = 10}

lymphoid_res <- FindAllMarkers(lymphoid,
                               assay = "escapeGSVA_normalized", 
                               return.thresh = 0.01,
                               logfc.threshold = 0.5,
                               min.pct = 0.20,
                               only.pos = T,
                               verbose = F)

lymphoid_res$pathway <- rownames(lymphoid_res)

lymphoid_res_sig <- lymphoid_res %>% filter(abs(avg_log2FC) > 0.1,
                               p_val_adj < 0.05)

plot_features <- lymphoid_res_sig %>%  group_by(cluster) %>% slice_max(order_by = abs(avg_log2FC), n = 20)

# Subset for heatmap
lymphoid_subset <- subset(lymphoid, downsample = 30000/length(unique(lymphoid$celltype)))

lymphoid_subset <- ScaleData(lymphoid_subset,
                             assay = "escapeGSVA_normalized")

DoHeatmap(
  lymphoid_subset,
  features = unique(plot_features$gene),
  group.by = "Treatment_Day",
  group.bar = TRUE,
  assay = "escapeGSVA_normalized")

```
